### 2023-06-30 reports/paper/paper.qmd
update paper

**Manuscript title:** Hypothetical limits on workplace exposure for
reducing non-Hodgkin lymphoma: An application of the hazard-extended
iterative conditional expectation parametric g-formula

@@

@@

**Running head suggestion:** Limits on metalworking fluid exposure to
prevent NHL.

@@

**Keywords:** Metalworking fluid, occupational exposure, non-Hodgkin
lymphoma, cancer, causal inference, g-formula, healthy-worker effect,
United States

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased
substantially in the US since 1960 and was recently linked to workplace
exposure to soluble metalworking fluids (MWF) in a standard survival
analysis of the United Auto Workers-General Motors (UAW-GM) cohort. To
further explore this association, we estimate the effects of
hypothetical limits on soluble MWF exposure in relation to NHL risk in
the same cohort of autoworkers.

**Methods:** We estimated counterfactual risk of NHL between 1985 and
2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n =
33,134) under hypothetical target exposure limits for soluble MWF. We
applied the hazard-extended iterative conditional expectation parametric
g-formula, an estimator capable of adjusting for time-varying
confounding affected by past exposure.

**Results:** 339 NHL cases occurred over the 30-year follow-up period.
Stronger target limits on soluble MWF exposure resulted in monotonic
reductions in NHL risk. Setting the target exposure limit at 2 mg/m^3^
would have prevented 98 (95% CI: 45, 161) cases of NHL. A target
exposure limit of 0.5 mg/m^3^, the NIOSH recommended exposure limit,
would have prevented 112 (95% CI: 57, 181) cases. Finally, a target
exposure limit of 0.05 mg/m^3^ would have prevented 129 (95% CI: 66,
214) cases.

**Conclusion:** Limits on soluble MWF exposure would have prevented
cases of NHL in the UAW-GM cohort. Although MWF use is expected to
decline as the automotive industry turns to electric vehicles, a boom in
domestic vehicle manufacturing could nonetheless lead to global
increases in worker exposure to MWF in the US. Maintaining or
strengthening exposure limits for MWF may protect workers from NHL.

@@

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled
between 1973 and 1994, before plateauing at around 19 per 100,000
persons per year, making it the seventh most common cancer in the
country [@seer_1994; @Ekstrom-Smedby_2006]. The strongest known risk
factor of NHL is immunosuppression, both congenital and acquired
[@Filipovich_1992; @Chiu_2015]. Modulation of immune function
contributes to the pathogenesis of NHL directly by perturbing lymphocyte
function or indirectly by reducing immunosurveillance against neoplasms.
Immunodeficiency may also interact with infection with Epstein-Barr
Virus, resulting in uncontrolled B-cell proliferation [@Oertel_2002].
Acquired immunodeficiency through infection with human immunodeficiency
virus (HIV) can lead to genetic lesions and dysregulation while
increasing vulnerability to infection by other oncogenic agents
[@Grulich_1999; @Alexander_2007]. However, since immunodeficiency and
infection with HIV are rare, they cannot fully explain the historic rise
or present burden of NHL [@Shiels_2013]. The rise in NHL incidence
coincided with a period of rapid and extensive chemicalization in
industry, agriculture and warfare; environmental and occupational
exposures may play an important explanatory role in the epidemiology of
NHL [@Nelson_2005; @Romero_2021].

@@

Metalworking fluids are complex mixtures of oil, water, and chemical
additives that optimize metal machining operations. During shaping,
grinding, and cutting operations, MWFs are misted, poured, or blasted at
high pressure onto work surfaces to remove debris, cool metal, improve
efficiency, and prevent deterioration of tools. Although MWFs are
essential to manufacturing processes, they also present a potential
health hazard to exposed workers through inhalation or ingestion of MWF
particulate mass.

There are three general types of MWF: straight, soluble, and synthetic.
The type of MWF applied has changed over the decades in response to
engineering, economic, and health considerations. Straight MWF, the
oldest class of MWF, is comprised mostly of mineral oils, which were
classified as carcinogenic by the International Agency for Research on
Cancer (IARC) in 1973, citing studies of occupational exposures among
workers in oil, textile, and metal industries [@IARC_1973]. In the same
decade, additional factors including the high cost of oil and the
introduction of performance-enhancing water-soluble chemical additives
spurred the gradual replacement of straight MWF with water-based soluble
and synthetic MWF for many, but not all metalworking operations
[@Childers_2005]. Although water-based MWFs contain little to no mineral
oil, they contain a richer cocktail of additives including chromates,
cycloalkanes, phenols, organochlorines, nitrites, sulfonates, and
triazines: classes of chemicals with known or plausible carcinogenicity
[@IARC_1987; @IARC_2014; @IARC_2017]. As the US prepares for a boom in
domestic automotive manufacturing, occupational exposure to MWF will
remain an important concern for worker health [@Armbrust_2021;
@Carr_2022; @Keilman_2023].

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects leave work [@Arrighi_1994]. A well-known analytical
method for causal inference in observational studies where the HWSE or
other forms of time-varying confounding affected by past exposure may be
operating is the parametric g-formula [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal
interpretation of any analysis is positivity (overlap) ie adequate
variation in the exposure of interest within strata formed by confounder
and exposure history [@Petersen_2012]. Here, we specify hypothetical
interventions on exposure to soluble MWF in the United Auto
Workers-General Motors (UAW-GM) occupational cohort whose effects can be
estimated with guaranteed positivity. In particular, we estimate the
cumulative incidence of NHL between 1985 and 2015 under well-supported
(supportable) interventions by applying the the hazard-extended
iterative conditional expectation (ICE) parametric g-formula
[@Arrighi_1994; @Wen_2021].

@@

We estimated NHL cumulative incidence from 1985 to 2015 under
supportable interventions on average daily exposure to soluble MWF by
applying the hazard-extended ICE parametric g-formula estimator
[@Wen_2020]. First, we estimated the expected number of NHL cases that
we would observe if there were no censoring by competing risks. Then, we
contrasted this case count to that under supportable interventions based
on five hypothetical target exposure limits and no competing risks. The
five target exposure limits were (1) 2.0, (2) 1.0, (3) 0.5, (4) 0.25,
and (5) 0.05 mg/m^3^. The National Institute for Occupational Safety and
Health (NIOSH) recommended exposure limit (REL) for total particulate
mass (PM) composed of MWF is 0.5 mg/m^3^ [@niosh_1998]. For each target
exposure limit, period of follow-up, and unique combination of
confounder and exposure histories, we also defined a supportable
exposure limit, which was the maximum observed value of exposure at or
below the target exposure limit or no limit when all of the observed
values were above the target limit. For each target exposure limit, we
specified an intervention rule that reduced exposures above the
supportable exposure limit to that limit, but allowed exposures at or
below the limit to vary according to their natural distribution.
Applying this intervention rule to the observed distribution of exposure
produced the intervention distribution that defined the stochastic
dynamic intervention with guaranteed positivity.

@@

We used data from the UAW-GM cohort, which included all hourly workers
at three automobile manufacturing plants in Michigan who had worked at
least three years by 1985. Past papers provide detailed descriptions of
the cohort [@Eisen_1992; @Eisen_2001]. The large size of the study
population and rich time-varying, quantitative MWF exposure data provide
an opportunity to study this relatively rare cancer and evaluate
realistic interventions on MWF exposure in a longitudinal cohort
setting. The present study population (N = 33,134) was restricted to the
autoworkers who were at work in 1941 or not yet hired, missing no more
than half of their employment history, and still alive at the start of
follow-up. Autoworkers in the study population were followed for NHL
incidence from January 1, 1985 until NHL diagnosis, death, December 31,
2014 (30 years after the end of employment record availability), or the
oldest observed age at death (108 years), whichever came earlier.

## Outcome and covariates

We identified incident cancers in the UAW-GM cohort between 1985 and
2014 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants
1 and 2, located in the greater Detroit metropolitan area, were also
linked to the Detroit Regional Registry of the Surveillance,
Epidemiology, and End Results (SEER) Program. Cancer types were
distinguished using site and histology codes conforming to the
International classification of Diseases for Oncology, 3^rd^ edition
(ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the
following ICD-O-3 histology codes: 9590-9597, 9670-9671, 9673, 9675,
9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712,
9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837.
Details regarding cancer incidence follow-up are described elsewhere
[@Colbeth_2023]. Vital status was ascertained from company records and
by linkage to Social Security Administration, National Death Index, and
state mortality files.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
occupational cohort studies. Exposure assessment was based on direct air
sampling as well as company records. Company industrial hygienists
collected several hundred personal and area samples for total
particulate matter (mg/m^3^) composed of MWF over many decades. Research
industrial hygienists collected additional air sampling data when the
cohort study was launched in the mid 1980s. These additional data
combined with historical data and company records constituted a
job-exposure matrix of quantitative 8-hour time-weighted average daily
exposure estimates to soluble, straight, and synthetic MWF for each
combination of job, department, and plant over time. Workers' average
daily exposure to each MWF type was determined by combining this
job-exposure matrix with employment records, which recorded time-varying
job type, department, and plant for each employee from hire to
termination or 1995, whichever came sooner. For employment records that
were at least half complete, gaps in the record were interpolated by
carrying forward the last known job type. The exposure assessment is
described in detail elsewhere [@Hallock_1994; @Woskie_1994;
@Woskie_2003].

Previous analyses of NHL applied exposure lags of 1 to 20 years to
account for disease latency; we lagged cumulative MWF exposures by 20
years and therefore ended follow-up on January 1, 2015 [@Smith_2007;
@Karipidis_2007; @Zhang_2019]. In analyses, MWF exposure history at
start of follow-up was summarized as the cumulative sum of annual
exposure intensities. Exposure was coded as categorical variables with
cut-points at zero and the quintiles of nonzero exposure among cases. We
estimated the effects of interventions on soluble MWF, the type of MWF
used most widely and in the greatest quantities. We treated exposure to
straight and synthetic MWF as potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the
deciles of the dates of diagnosis of NHL. The number of years per period
ranged from two to four years. The hazard-extended ICE parametric
g-formula involved two stages. In the first, we estimated counterfactual
discrete hazards over the person-periods. In the second, we pooled those
estimates to estimate the counterfactual risk over the entire follow-up
period. During pooling, we iteratively combined estimates of the
counterfactual hazard starting with the last two periods, then the last
three, then the last four, and so forth. In each iteration, we performed
model-based standardization over exposure and covariate histories before
combining the counterfactual discrete hazard with the risk over
subsequent periods. This iterative process resulted in a sequentially
standardized estimate of the counterfactual cumulative incidence of NHL
when the intervention of interest is enforced over all follow-up
periods. We took exposure to straight and synthetic MWF, employment
status, cumulative time off, age, duration of employment, sex
(male/female), race (Black/white/unknown), and plant (Plant 1/Plant
2/Plant 3) to be potential confounders. Exposure to MWF, employment
status, time off, and duration of employment were lagged 20 years. An
overview of the general steps of the estimation procedure are presented
below. Note that we refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression of NHL on covariates and exposure
    over all at-risk person-periods.

@@

3.  Within strata formed by unique combinations of covariates, obtain
    the intervention distribution of exposure by applying the
    intervention rule to the observed exposure distribution.

@@

    a)  Regress the counterfactual hazard pooled over all subsequent
        periods on exposure and covariate history at the current period.
    b)  Predict the pooled hazard for each person at risk at the start
        of the period and each possible level of exposure using the
        model from (a). Retrieve the predicted hazards from step 2. For
        each person and level of exposure, multiply the predicted pooled
        hazard by one minus the corresponding predicted hazard and add
        the predicted hazard to the product.
    c)  Within strata formed by exposure and covariate history, obtain
        the intervention distribution of exposure by applying the
        intervention rule to the observed exposure distribution.
    d)  Within strata, estimate the counterfactual hazard pooled over
        the present and subsequent periods for each person by taking a
        weighted sum of the predicted pooled hazards from (b).
    e)  If the present period is not first period, set the reference
        period to the preceding period and return to (a).

6.  Compute the counterfactual risk by averaging the counterfactual
    hazard over the entire follow-up period across all persons.

@@

We estimated cumulative incidence under the observed distribution of
soluble MWF exposure (natural course) and under the five supportable
exposure limits. We contrasted the cumulative incidence under
supportable limits to that under the natural course by computing the
number of cases averted and the cumulative incidence ratios. Confidence
intervals were computed using the basic nonparametric bootstrap with
1000 Monte Carlo samples from the population at the start of follow-up.
All the necessary script used to reproduce the analyses are available on
[GitHub](https://github.com/kvntchn/gm-nhl-ice).

```{r}
source(here::here("../gm-wrangling/wrangling/table1.R"))
clean.tab1 <- function(tab1) {
  # hspace.which <- grepl("\\\\hspace\\{.*\\}", rownames(tab1))
  rownames(tab1) <- gsub("\\\\hspace\\{.*\\}|\\\\hline |\\$.*\\$| among deceased", "", rownames(tab1))

@@

Table 1 presents summary statistics of exposure and covariates for the
full study population and for those diagnosed with NHL between 1985 and
2005. The cohort is predominantly white (66%) and male (87%). The median
year of hire among those diagnosed with NHL was 1959 whereas the median
year of hire in the full study population was almost a decade later. Age
at hire was approximately the same among those with NHL and the full
study population. Median lagged cumulative exposure to all three MWF
types was higher among NHL cases. Soluble MWF were the most widely used
MWF type, with approximately 90% of workers ever exposed. Median
cumulative exposure among the exposed was 6.5 times higher for soluble
than for straight MWF. Figure 1 shows median annual average daily
exposure to the three MWF types among exposed workers over calendar
time. Exposure to MWF generally followed a downward trend over time.

```{r}
# tab1 <- tab1[tab1.which,]
tab1[c(1:13, 15, 14, 16:27, 29, 28, 30),-grep("hspace.logical", colnames(tab1))] |> 
  data.frame() |> flextable() |>
  delete_part(part = "header") |>
  add_header_row(values = c("", "Study population", "", "NHL cases"),
                 colwidths = c(1, 2, 1, 2)) |>
  hline_top(part = "header") |>
  hline_top() |> hline_bottom() |>
  hline(which((grepl("[0-9]\\)", tab1$spread))[-1])[1]) |> width(4, 0.1) |>
  width(1, 2.2) |> width(c(3, 6), 1.1) |> width(c(2, 5), 0.65) |>
  flextable::compose(
    i = grep("Cumulative exposure", rownames(tab1)), j = 1,
    value = as_paragraph("Cumulative exposure to MWF (mg/m",
                         as_sup("3"), "-years)")) |>
  flextable::align(j = 2:6, align = "center", part = "all") |>
  footnote(i = grep("Plant$", rownames(tab1)), j = 1, value = as_paragraph(
    " Plant of longest employment duration among those who worked at multiple plants"),
    ref_symbols = "a", inline = T) |>
  footnote(i = grep("Ever exposed|time off", rownames(tab1)), j = 1, value = as_paragraph(
    " Lagged 10 years"),
    ref_symbols = "b", inline = T) |>
  footnote(i = grep("work", rownames(tab1)), j = 1, value = as_paragraph(
    " Among those who left work by December 31, 1994"),
    ref_symbols = "c", inline = T) |>
  # footnote(i = grep("death", rownames(tab1)), j = 1, value = as_paragraph(
  #   " Among deceased"),
  #   ref_symbols = "d", inline = T) |>
  footnote(i = grep("Cumulative exposure", rownames(tab1)), j = 1, value = as_paragraph(
    " Among ever-exposed individuals, lagged 10 years."),
    ref_symbols = "d", inline = T) |>
  add_footer_row(values = "NHL: non-Hodgkin lymphoma.", colwidths = 6) |>
  padding(padding = 1.5, part = "all") |>
  padding(
    i = which(tab1$hspace.logical),
    j = 1,
    padding.left = 20
  ) |>
  merge_at(i = grep("Cumulative exposure", rownames(tab1))) |>
  hline_bottom(part = "footer") |> font(fontname = 'arial') |>
  set_caption(caption = "Summary of population characteristics. Statistics shown above the horizontal line are count (%). Those shown below are median (quartile 1, quartile 3).")
```

@@

![Figure 1. Median annual average daily exposure to soluble, straight,
and synthetic metalworking fluids among exposed workers over
time.](../../resources/images/exposure.png){width="6.5in"}

@@

The observed number of NHL cases over the 30-year follow-up period was
339. Table 2 presents the hazard-extended ICE parametric g-formula
estimates of the counterfactual number of cases, number of cases
averted, and cumulative incidence ratios contrasting hypothetical
interventions on exposure to soluble MWF based on different target
exposure limits and no competing risks. Under an intervention
eliminating competing risks, the estimated number of cases was 502 (95%
CI: 439, 555). Interventions based on stronger limits on average daily
exposure to soluble MWF resulted in monotonically stronger reductions in
the estimated cumulative incidence of NHL. Setting the target exposure
limit at the NIOSH REL 0.5 mg/m^3^ would have averted 112 (95% CI: 57,
181) NHL cases.

>>>>

**Manuscript title:** Hypothetical interventions on workplace exposure
with guaranteed positivity

@@

^b^ University of California, Berkeley School of Public Health, Division
of Environmental Health Sciences

@@

**Running head suggestion:** Supportable interventions for reducing NHL
risk.

@@

**Keywords:** Metalworking fluid, occupational exposure, non-Hodgkin
lymphoma, cancer, positivity, overlap, causal inference, g-formula,
iterated conditional expectation, healthy-worker effect

**Word count** **(Limit: 1500 to 4000 words):** 3812

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased
substantially in the US since 1960 and was recently linked to workplace
exposure to soluble metalworking fluids (MWF) in a standard survival
analysis of the United Auto Workers-General Motors (UAW-GM) cohort.
Here, we estimate the effects of hypothetical interventions on soluble
MWF exposure in relation to NHL risk in the same cohort of autoworkers.
Estimation of causal effects requires positivity, which we guarantee by
investigating causal contrasts that are supportable by the data.

**Methods:** We estimated counterfactual risk of NHL between 1985 and
2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n =
33,134) under supportable interventions on exposure to soluble MWF. The
supportable intervention reduces annual daily average exposures above
the target limit to that limit when the target limit is supported by
data. Otherwise, the supportable intervention reduces exposure to the
highest level of exposure below the target limit or not at all,
depending on support.

**Results:** 339 NHL cases occurred over the 30-year follow-up period.
Stronger target limits on soluble MWF exposure resulted in monotonic
reductions in NHL risk. Setting the target exposure limit at 0.5
mg/m^3^, the NIOSH recommended exposure limit, would have prevented 112
(95% CI: 57, 181) cases.

**Conclusion:** Stronger limits on exposure to soluble MWF provide
stronger protections against NHL. We expect the effects of enforcing the
target exposure limits uniformly to have even stronger protective
effects. Strengthening exposure limits for MWF may protect workers from
NHL during the anticipated boom in domestic manufacturing.

@@

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled
between 1973 and 1994, before plateauing at around 19 per 100,000
persons per year, making it the seventh most common cancer in the
country [@seer_1994; @Ekstrom-Smedby_2006]. The strongest known risk
factor of NHL is immunosuppression, both congenital and acquired
[@Filipovich_1992; @Chiu_2015]. However, since immunodeficiency and
infection with HIV are rare, they cannot fully explain the historic rise
or present burden of NHL [@Shiels_2013]. The rise in NHL incidence
coincided with a period of rapid and extensive chemicalization in
industry, agriculture and warfare; environmental and occupational
exposures may play an important explanatory role in the epidemiology of
NHL [@Nelson_2005; @Romero_2021].

@@

Metalworking fluids are complex mixtures of oil, water, and chemical
additives that optimize metal machining operations. There are three
general types of MWF: straight, soluble, and synthetic. During shaping,
grinding, and cutting operations, MWFs are misted, poured, or blasted at
high pressure onto work surfaces to remove debris, cool metal, improve
efficiency, and prevent deterioration of tools. Although MWFs are
essential to manufacturing processes, they also present a potential
health hazard to exposed workers through inhalation or ingestion of MWF
particulate mass. In response to health concerns and the rising cost of
oil, soluble MWF were developed to replace straight MWF and are now the
most commonly used type [@Byers_2006].

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects of exposure leave work [@Arrighi_1994]. An early causal
inference method in statistics developed to estimate causal effects in
longitudinal observational studies where the HWSE or other forms of
time-varying confounding/selection bias affected by past exposure may be
operating is the parametric g-formula [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal inference
is positivity (overlap) ie adequate variation in the exposure of
interest within strata formed by confounder and exposure histories
[@Petersen_2012].

Common approaches for addressing violations in positivity are
model-based smoothing when using outcome-regression techniques, weight
truncation when using propensity methods, or a combination of both.
Another approach is to consider only those causal contrasts that are
supported by the observed data ie have strictly positive propensity
score. Here, we defined supportable interventions on exposure to soluble
MWF in the United Auto Workers-General Motors (UAW-GM) occupational
cohort whose effects can be estimated with guaranteed positivity. We
estimated the cumulative incidence of NHL between 1985 and 2015 under
supportable interventions based on selected target exposure limits on
annual average daily exposure by applying the the hazard-extended
iterative conditional expectation (ICE) parametric g-formula
[@Arrighi_1994; @Wen_2020].

@@

We estimated NHL cumulative incidence from 1985 to 2015 under
supportable interventions on on annual average daily exposure to soluble
MWF based on several target exposure limits. First, we estimated the
expected number of NHL cases that we would observe if there were no
censoring by censoring and no target exposure limit. Then, we contrasted
this case count to that under supportable interventions based on five
hypothetical target exposure limits and no censoring. The five target
exposure limits were (1) 2.0, (2) 1.0, (3) 0.5, (4) 0.25, and (5) 0.05
mg/m^3^. The National Institute for Occupational Safety and Health
(NIOSH) recommended exposure limit (REL) for time-weighted average total
particulate mass (PM) composed of MWF is 0.5 mg/m^3^ [@niosh_1998].

For each target exposure limit, period of follow-up, and strata defined
by unique combinations of confounder and exposure histories, we found a
supportable exposure limit, which was the maximum observed value of
exposure at or below the target exposure limit. If all of the observed
values were above the target limit, no limit was enforced for that
stratum. The supportable intervention rule then reduces exposures above
the supportable exposure limit to that limit, but allowed exposures at
or below the limit to vary according to their observed distribution.
Applying the supportable intervention rule to the observed distribution
of exposure produces the intervention distribution that defines the
corresponding stochastic dynamic intervention with guaranteed
positivity. We estimated the effect of supportable intervention rules
based on selected target exposure limits, expressed as stochastic
dynamic interventions, using the hazard-extended ICE parametric
g-formula [@Wen_2020].

Figure 1 presents three example scenarios where the target exposure
limit is 0.25 mg/m^3^ in all cases, but the supportable exposure limit
differs depending on what the observed data supports. In Figure 1a, the
supportable exposure limit is equal to the target exposure limit as some
individuals with that particular set of potential confounder and
exposure histories were observed to have experienced exposure at that
level. In Figure 1b, the supportable exposure limit is below the target
exposure limit. In Figure 1c, there is no limit on exposure because no
individuals in that stratum experienced exposure below the target
exposure limit.

::: {#fig-rules layout-ncol="2"}
![](resources/images/intervention-density-witnin-combos-a.png){#fig-1a
width="3.5in"}
![](resources/images/intervention-density-witnin-combos-b.png){#fig-1b
width="3.5in"}
![](resources/images/intervention-density-witnin-combos-c.png){#fig-1c
width="3.5in"}

Observed and post intervention distribution of nonzero exposure for
three distinct confounder and exposure histories before and after
applying the supportable intervention rule.
:::

@@

We used data from the UAW-GM cohort, which included all hourly workers
at three automobile manufacturing plants in Michigan who had worked at
least three years by 1985. Past papers provide detailed descriptions of
the cohort [@Eisen_1992; @Eisen_2001]. The large size of the study
population and rich time-varying, quantitative MWF exposure data enable
the study of this relatively rare cancer and evaluate realistic
interventions on MWF exposure in a longitudinal cohort setting. The
present study population (N = 33,134) was restricted to the autoworkers
who were at work in 1941 or not yet hired, missing no more than half of
their employment history, and still alive at the start of follow-up.
Autoworkers in the study population were followed for NHL incidence from
January 1, 1985 until NHL diagnosis, death, December 31, 2014 (30 years
after the end of employment record availability), or the oldest observed
age at death (108 years), whichever came earlier.

## Outcome and potential confounders

We identified incident cancers in the UAW-GM cohort between 1985 and
2014 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants
1 and 2, located in the greater Detroit metropolitan area, were also
linked to the Detroit Regional Registry of the Surveillance,
Epidemiology, and End Results (SEER) Program. Cancer types were
distinguished using site and histology codes conforming to the
International classification of Diseases for Oncology, 3^rd^ edition
(ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the
following ICD-O-3 histology codes: 9590-9597, 9670-9671, 9673, 9675,
9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712,
9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837.
Details regarding cancer incidence follow-up are described elsewhere
[@Colbeth_2022]. Vital status was ascertained from company records and
by linkage to Social Security Administration, National Death Index, and
state mortality files.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
occupational cohort studies. Exposure assessment was based on direct air
sampling as well as company records. Company industrial hygienists
collected several hundred personal and area samples for total
particulate matter (mg/m^3^) composed of MWF over many decades. Research
industrial hygienists collected additional air sampling data when the
cohort study was launched in the mid 1980s. These additional data
combined with historical data and company records constituted a
job-exposure matrix of quantitative 8-hour time-weighted average daily
exposure estimates to soluble, straight, and synthetic MWF for each
combination of job, department, and plant over time. Workers' annual
average daily exposure to each MWF type was determined by combining this
job-exposure matrix with employment records, which recorded time-varying
job type, department, and plant for each employee from hire to
termination or 1995, whichever came sooner. For employment records that
were at least half complete, gaps in the record were interpolated by
carrying forward the last known job type. The exposure assessment is
described in detail elsewhere [@Hallock_1994; @Woskie_1994;
@Woskie_2003].

Previous analyses of NHL applied exposure lags of 1 to 20 years to
account for disease latency; we lagged cumulative MWF exposures by 20
years and therefore ended follow-up on January 1, 2015 [@Smith_2007;
@Karipidis_2007; @Zhang_2019]. In analyses, MWF exposure history at
start of follow-up was summarized as the cumulative sum of average
exposure intensities. Exposure was coded as categorical variables with
cut-points at zero and the quintiles of nonzero exposure among cases. We
estimated the effects of interventions on soluble MWF, the type of MWF
used most widely and in the greatest quantities while treating
co-exposure to straight and synthetic MWFs as potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the
deciles of the dates of diagnosis of NHL. The number of years per period
ranged from two to four years. The hazard-extended ICE parametric
g-formula involves two stages. In the first, we estimate counterfactual
discrete hazards over the person-periods. In the second, we pool those
estimates to estimate the counterfactual risk over the entire follow-up
period. During pooling, we iteratively combine estimates of the
counterfactual hazard to obtain a pooled estimate over an increasing
number of periods starting from the last period. In each iteration, we
perform model-based standardization over exposure and covariate
histories before combining the counterfactual discrete hazard estimate
with the estimated hazard pooled over subsequent periods. This iterative
process results in a sequentially standardized estimate of the
counterfactual cumulative incidence of NHL when the intervention of
interest is enforced over all follow-up periods.

We investigated supportable interventions that guarantee positivity:
every value of exposure that could be assigned under the our stochastic
dynamic intervention has a strictly positive propensity score. We took
exposure to straight and synthetic MWF, employment status, cumulative
time off, age, duration of employment, sex (male/female), race
(Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3) to be
potential confounders. Exposure to MWF, employment status, time off, and
duration of employment were lagged 20 years. An overview of the general
steps of the estimation procedure are presented below. Note that we
refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression of NHL on potential confounders and
    exposure over all at-risk person-periods.

@@

3.  Within strata formed by unique combinations of potential confounders
    and exposure, obtain the intervention distribution of exposure by
    applying the supportable intervention rule to the observed exposure
    distribution.

@@

    a)  Regress the counterfactual hazard pooled over all subsequent
        periods on all past potential confounders and exposures.
    b)  Predict the pooled hazard for each person at risk at the start
        of the period and each possible level of exposure using the
        model from (a). Retrieve the predicted hazards from step 2. For
        each person and level of exposure, multiply the predicted pooled
        hazard by one minus the corresponding predicted hazard. Then,
        add the predicted hazard to the product.
    c)  Within strata formed by potential confounder and exposure
        histories, obtain the intervention distribution of exposure by
        applying the intervention rule to the observed exposure
        distribution.
    d)  Within strata, estimate the counterfactual hazard pooled over
        the present and subsequent periods for each person by taking a
        weighted sum of the predicted pooled hazards from (b).
    e)  If the present period is not first period, set the reference
        period to the preceding period and return to (a).

6.  Compute the counterfactual risk by averaging the pooled
    counterfactual hazards across all persons.

@@

We estimated cumulative incidence under no intervention on soluble MWF
and under the supportable interventions based on the five target
exposure limits. We contrasted the cumulative incidence under the
supportable interventions to that under no intervention on exposure by
computing the number of cases averted and the cumulative incidence
ratios. Confidence intervals were computed using the basic nonparametric
bootstrap with 1000 Monte Carlo samples from the population at the start
of follow-up. All the necessary script used to reproduce the analyses
are available on [GitHub](https://github.com/kvntchn/gm-nhl-ice).

```{r}
source(here::here("../gm-wrangling/wrangling/table1.R"))
clean.tab1 <- function(tab1) {
  # hspace.which <- grepl("\\\\hspace\\{.*\\}", rownames(tab1))
	rownames(tab1) <- gsub(
		"\\\\hspace\\{.*\\}|\\\\hline |\\$.*\\$| among deceased",
		"",
		rownames(tab1))

@@

Table 1 presents summary statistics for the full study population and
among those diagnosed with NHL between 1985 and 2015. The cohort is
predominantly white (64%) and male (86%). Year of hire and age at hire
were approximately the same among those with NHL and the full study
population. Median lagged cumulative exposure to all three types of MWF
was higher among NHL cases. Soluble MWF were the most widely used type,
with approximately 88% of workers ever exposed. Median cumulative
exposure among the exposed was 6 times higher for soluble than for
straight MWF. Figure 2 shows median annual average daily exposure to the
three MWF types among exposed workers over calendar time. Exposure to
MWF generally followed a downward trend over time.

```{r}
# tab1 <- tab1[tab1.which,]
tab1.sub <- tab1[c(1:13, 15, 14, 16:27, 29, 28, 30),]
tab1.sub <- tab1.sub[-c(17, 21, 26),]
tab1.sub_hspace <- tab1.sub$hspace.logical
tab1.sub <- tab1.sub[,-grep('hspace.logical', names(tab1.sub))]
```

```{r}
tab1.sub |> 
  data.frame() |> flextable() |>
  delete_part(part = "header") |>
  add_header_row(values = c("", "Study population", "", "NHL cases"),
                 colwidths = c(1, 2, 1, 2)) |>
  hline_top(part = "header") |>
  hline_top() |> hline_bottom() |>
  hline(which((grepl("[0-9]\\)", tab1.sub$spread))[-1])[1]) |> width(4, 0.1) |>
  width(1, 2.2) |> width(c(3, 6), 1.1) |> width(c(2, 5), 0.65) |>
  flextable::compose(
    i = grep("Cumulative exposure", rownames(tab1.sub)), j = 1,
    value = as_paragraph("Cumulative exposure to MWF (mg/m",
                         as_sup("3"), "-years)")) |>
  flextable::align(j = 2:6, align = "center", part = "all") |>
  footnote(i = grep("Plant$", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Plant of longest employment duration among those who worked at multiple plants"),
    ref_symbols = "a", inline = T) |>
  footnote(i = grep("Ever exposed|time off", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Lagged 10 years"),
    ref_symbols = "b", inline = T) |>
  footnote(i = grep("work", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Among those who left work by December 31, 1994"),
    ref_symbols = "c", inline = T) |>
  # footnote(i = grep("death", rownames(tab1.sub)), j = 1, value = as_paragraph(
  #   " Among deceased"),
  #   ref_symbols = "d", inline = T) |>
  footnote(i = grep("Cumulative exposure", rownames(tab1.sub)), j = 1, value = as_paragraph(
    " Among ever-exposed individuals, lagged 10 years."),
    ref_symbols = "d", inline = T) |>
  add_footer_row(values = "NHL: non-Hodgkin lymphoma.", colwidths = 6) |>
  padding(padding = 1.5, part = "all") |>
  padding(
    i = which(tab1.sub_hspace),
    j = 1,
    padding.left = 20
  ) |>
  merge_at(i = grep("Cumulative exposure", rownames(tab1.sub))) |>
  hline_bottom(part = "footer") |> font(fontname = 'arial') |>
	set_caption(
		caption = paste0(
			"Summary of population characteristics. Statistics shown above ",
			"the horizontal line are count (%). Those shown below are median ",
			"(quartile 1, quartile 3).")
	)
```

@@

![Figure 2. Median annual average daily exposure to soluble, straight,
and synthetic metalworking fluids among exposed workers over
time.](../../resources/images/exposure.png){width="6.5in"}

@@

The observed number of NHL cases over the 30-year follow-up period was
339. Table 2 presents the hazard-extended ICE parametric g-formula
estimates of the counterfactual number of cases, number of cases
averted, and cumulative incidence ratios contrasting supportable
interventions on exposure to soluble MWF based on different target
exposure limits and no censoring. Under an intervention eliminating
censoring, the estimated number of cases was 502 (95% CI: 439, 555).
Interventions based on stronger target exposure limits on soluble MWF
resulted in monotonically stronger reductions in the estimated
cumulative incidence of NHL. Setting the target exposure limit at the
NIOSH REL 0.5 mg/m^3^ would have averted 112 (95% CI: 57, 181) NHL
cases.
